\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-08 k-planar.tex
Small changes, include K -> T

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example.  
So, loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded 
\begin{wrapfigure}{r}{0.5\textwidth} 
\centering
\includegraphics{ProductExample}
\caption{Example of a strong product.}
\label{ProductExample}
\end{wrapfigure}
treewidth to be generalised for planar graphs (with different constants). \cref{kPlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} proved that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}), and \citet{dujmovic.esperet.ea:planar}  proved that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).

@@

\begin{obs}
\label{AddDummy}
Every $k$-planar graph is isomorphic to $G^\PP$ for some planar graph $G$ and some $(k+1,2)$-shortcut system $\PP$ for $G$. 
\end{obs}

@@

\begin{thm}
\label{kPowerBasic}
For every planar graph $G$ with maximum degree $\Delta$ and for every integer $k\geq 1$, $G^k$ is a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth at most $O( k^7 \Delta^{k})$ and for some path $P$. \note{DW}{Improve to $k^6\Delta^{k}$?}
\end{thm}

@@

We now define two concepts used in the statements of \cref{PlanarBasic,kPlanarBasic}: strong products and treewidth. The \emph{strong product} of graphs $A$ and $B$, denoted by $A\boxtimes B$, is the graph with vertex set $V(A)\times V(B)$, where distinct vertices $(v,x),(w,y)\in V(A)\times V(B)$ are adjacent if: 
\begin{compactitem}
\item  $v=w$ and $xy\in E(B)$, or 
\item  $x=y$ and $vw\in E(A)$, or  
\item  $vw\in E(A)$ and $xy\in E(B)$. 
\end{compactitem}
A \emph{tree-decomposition} $\mathcal{K}$ of a graph $G$ consists of a tree $K$ and a collection $\mathcal{K}=(B_x:x\in V(K))$ of subsets of $V(G)$ indexed by the nodes of $K$ such that:
\begin{compactenum}[(i)]
\item for every $vw\in E(G)$, there exists some $x\in V(K)$ with $v,w\in B_x$; and 
\item for every $v\in V(G)$, the induced subgraph $K[x] := K[\{y: x\in B_y\}]$ is connected.  
\end{compactenum}
The \emph{width} of the tree-decomposition $\mathcal{K}$ is $\max\{|B_x|:x\in V(K)\}-1$.  The \emph{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.  Treewidth is the standard measure of how similar a graph is to a tree. Indeed, a connected graph has treewidth 1 if and only if it is a tree. Treewidth is of fundamental importance in structural and algorithmic graph theory; see \citep{Reed03,HW17,Bodlaender-TCS98} for surveys.

@@

Prior to this work, the strongest structural description of $k$-planar or $(g,k)$-planar graphs was in terms of layered treewidth, which we now define.  A \emph{layered tree-decomposition} $(\mathcal{L},\mathcal{K})$ consists of a layering $\mathcal{L}$ and a tree-decomposition $\mathcal{K}$ of $G$. The layered width of $(\mathcal{L},\mathcal{K})$ is $\max\{|L\cap B|: L\in \mathcal{L},\, B\in \mathcal{K}\}$.  The \emph{layered treewidth} of $G$ is the minimum layered width of any layered tree-decomposition of $G$. \citet{dujmovic.morin.ea:layered} proved that planar graphs have layered treewidth at most 3, that graphs of Euler genus $g$ have layered treewidth at most $2g+3$, and more generally that a minor-closed class has bounded layered treewidth if and only if it excludes some apex graph. \citet{dujmovic.eppstein.ea:structure} show that every $k$-planar graph has layered treewidth at most $6(k+1)$, and more generally that 
every $(g,k)$-planar graph has layered treewidth at most $(4g+6)(k+1)$.

>>>>

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example.  
So, loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded 
\begin{wrapfigure}{r}{0.5\textwidth} 
\centering
\includegraphics{ProductExample}
\caption{Example of a strong product.}
\label{ProductExample}
\end{wrapfigure}
treewidth to be generalised for planar graphs (with different constants). \cref{kPlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} proved that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92} from 1992), and \citet{dujmovic.esperet.ea:planar}  proved that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02} from 2002).

@@

\begin{obs}
\label{AddDummy}
Every $k$-planar graph is isomorphic to a subgraph of $G^\PP$ for some planar graph $G$ and some $(k+1,2)$-shortcut system $\PP$ for $G$. 
\end{obs}

@@

\begin{thm}
\label{kPowerBasic}
For every planar graph $G$ with maximum degree $\Delta$ and for every integer $k\geq 1$, $G^k$ is a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth at most $O( k^7 \Delta^{k})$ and for some path $P$. \note{DW}{Improve to $k^6\Delta^{k}$?}
\end{thm}
\note{PM}{TODO: Work out the treewidth more carefully. The use of big-O is here difficult to decipher (and probably incorrect).}

@@

We now define two concepts used in the statements of \cref{PlanarBasic,kPlanarBasic}: strong products and treewidth. The \emph{strong product} of graphs $A$ and $B$, denoted by $A\boxtimes B$, is the graph with vertex set $V(A)\times V(B)$, where distinct vertices $(v,x),(w,y)\in V(A)\times V(B)$ are adjacent if: 
\begin{compactitem}
\item  $v=w$ and $xy\in E(B)$, or 
\item  $x=y$ and $vw\in E(A)$, or  
\item  $vw\in E(A)$ and $xy\in E(B)$. 
\end{compactitem}
A \emph{tree-decomposition} $\mathcal{T}$ of a graph $G$ consists of a tree $T$ and a collection $\mathcal{T}=(B_x:x\in V(T))$ of subsets of $V(G)$ indexed by the nodes of $K$ such that:
\begin{compactenum}[(i)]
\item for every $vw\in E(G)$, there exists some node $x\in V(T)$ with $v,w\in B_x$; and 
\item for every $v\in V(G)$, the induced subgraph $T[v] := T[\{x: v\in B_x\}]$ is connected.  
\end{compactenum}
The \emph{width} of the tree-decomposition $\mathcal{T}$ is $\max\{|B_x|:x\in V(T)\}-1$.  The \emph{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.  Treewidth is the standard measure of how similar a graph is to a tree. Indeed, a connected graph has treewidth 1 if and only if it is a tree. Treewidth is of fundamental importance in structural and algorithmic graph theory; see \citep{Reed03,HW17,Bodlaender-TCS98} for surveys.

@@

Prior to this work, the strongest structural description of $k$-planar or $(g,k)$-planar graphs was in terms of layered treewidth, which we now define.  A \emph{layered tree-decomposition} $(\mathcal{L},\mathcal{T})$ consists of a layering $\mathcal{L}$ and a tree-decomposition $\mathcal{T}$ of $G$. The layered width of $(\mathcal{L},\mathcal{T})$ is $\max\{|L\cap B|: L\in \mathcal{L},\, B\in \mathcal{T}\}$.  The \emph{layered treewidth} of $G$ is the minimum layered width of any layered tree-decomposition of $G$. \citet{dujmovic.morin.ea:layered} proved that planar graphs have layered treewidth at most 3, that graphs of Euler genus $g$ have layered treewidth at most $2g+3$, and more generally that a minor-closed class has bounded layered treewidth if and only if it excludes some apex graph. \citet{dujmovic.eppstein.ea:structure} show that every $k$-planar graph has layered treewidth at most $6(k+1)$, and more generally that 
every $(g,k)$-planar graph has layered treewidth at most $(4g+6)(k+1)$.
